miniKanren with fair search strategies

### 2019-05-19 Paper/paper.tex
minor improvements

The syntax of a programming language should reflect its semantics. When writing 
a \conde{} expression in miniKanren, a programmer would expect all clauses 
share the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving \DFSi{}, 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \conde{} is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by \DFSi{} is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj{} and \conj{}.

@@

The miniKanren in TR2 implements interleaving DFS(\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

\DFSbi{} is fair only when the number of goals is a power of 2, 
otherwise, some goals would be allocated twice as many resources than the 
others. In the above example, where the \conde has five clauses, the clauses of 
\texttt{b}, \texttt{c}, and \texttt{d} are allocated more resources.

@@

Under the hood, the \conde{} and the call to \texttt{repeato} are connected by 
\conj{}. The \conde{} goal outputs a search space including four trivially 
different states. These states are then supplied individually to 
\texttt{(repeato x q)} goal, producing four search spaces. In the examples 
above, the \conj{}s are allocating more computational resources to the state of 
\texttt{a}. On the contrary, fair \conj{} would allocate resources evenly to 
each search space.

@@

A more interesting situation is when the first conjunct produces infinite many 
answers. Consider the following example, a naive specification of fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but no 
longer ones, which makes the strategies incomplete. A search strategy is 
\emph{complete} iff it can find out all the answers within finite time, 
otherwise, 
it is \emph{imcomplete}.

@@

BFS puts states of the same cost in the same bag, and allocate resources 
carefully among search spaces related to different bags such that answers are 
produced in increasing order of cost. The \emph{cost} of a answer is its depth 
in the search tree (i.e. the number of calls to relational definitions required 
to find them) \citet{seres1999algebra}. In the following example, the cost of 
each answer is equal to the length of the inner lists plus the length of the 
outer list.

@@

\begin{defn}[fair \conj{}]
A \conj{} is fair iff it allocates computational resource evenly to 
search spaces produced from states in the same bag. A bag is a finite 
collection of state. And search strategies with fair \conj{} should represent 
search spaces with possibly infinite collection of state. 
\end{defn}

\begin{defn}[unfair \conj{}]
A \conj{} is unfair iff it is not fair.
\end{defn}

@@

Fig.~\ref{DFSi} shows part of the implementation of \DFSi{}. We follow a
convention to name variables bound to states with `s', to name variables bound 
to goals with `g', and to name variables bound to search spaces with a suffix 
`-inf'. The first function, \texttt{disj2}, implements binary disjunction. 
\texttt{append-inf} is its helper, which compose two disjunctive search spaces. 
The following function, \texttt{conj2}, implements binary conjunction. It 
applies 
the \emph{first} goal to the input state, then applies the second goal to 
states in the resulting search space. The latter process is done with a helper 
function. \texttt{append-map-inf} applys its input goal to states in its input 
search spaces and compose the resulting search spaces. It reuse 
\texttt{append-inf} for search space composition. The following three 
definitions introduce syntactic sugars that miniKanren users are more familiar 
with. The first two definitions say disjunction and disjunction are 
right-associative. The next and last definition says \conde{} relates its 
clauses disjunctively, and goals in the same clause conjunctively.

@@

\begin{figure}
	\lstinputlisting{Figures/balanced-disj.rkt}
	\caption{DFS$_\textrm{bi}$ implementation}
	\label{balanced-disj}
\end{figure}

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair 
\conj{}. The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ 
in the \disj{} macro. We list the new \disj{} with its helper in 
Fig.~\ref{balanced-disj}. The helper \code{disj+} builds a balanced binary tree 
whose leaves are the goals and whose nodes are \code{disj2}, hence the name of 
this search strategy. The first argument 
to \code{disj+} is the goals. And the next two arguments accumulate goals in 
the left and right subtrees. The first clause says that when there is one goal, 
the tree is the goal itself. When there is more goals, they are partitioned 
into two sub-lists. The partition is done by repetitively dispatch the first 
two goals (the last clause), until no goal remains (the second clause) or one 
goal remains (the third clause). In contrast, the \disj{} in DFS$_\textrm{i}$ 
constructs the binary tree with the same collection of leaf nodes but in a 
particularly unbalanced form.

@@

\begin{figure}
	\lstinputlisting{Figures/fDFS.rkt}
	\caption{DFS$_\textrm{f}$ implementation}
	\label{fDFS}
\end{figure}

@@

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our BFS implementation is not an 
exception: \texttt{none}, \texttt{unit}, \texttt{append-map-inf}, and 
\texttt{append-inf} correspond to \texttt{mzero}, \texttt{unit}, \texttt{bind}, 
and \texttt{mplus} respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s-inf} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is an eliminator of search 
space, hence the name. The remaining functions do the same thing as before.

@@

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower. Details about 
difference in efficiency are in section 6.

@@

We devise three new search strategies: balanced interleaving DFS (DFS$_\textrm{bi}$), fair 
DFS (DFS$_\textrm{f}$), and BFS. DFS$_\textrm{bi}$ has almost-fair \disj{} and unfair \conj{}. DFS$_\textrm{f}$ has 
fair \disj{} and unfair \conj{}. BFS has both fair \disj{} and fair \conj{}.

>>>>

The syntax of a programming language should reflect its semantics. When writing 
a \conde{} expression in miniKanren, a programmer would expect all clauses 
share the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving depth-first search 
(\DFSi{}), 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \conde{} is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by \DFSi{} is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj{} and \conj{}.

@@

The miniKanren in TR2 implements interleaving DFS (\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

\DFSbi{} is fair only when the number of goals is a power of 2, 
otherwise, some goals would be allocated twice as many resources than the 
others. In the above example, where the \conde{} has five clauses, the clauses 
of \texttt{b}, \texttt{c}, and \texttt{d} are allocated more resources.

@@

Under the hood, the \conde{} and the call to \texttt{repeato} are connected by 
\conj{}. The \conde{} goal outputs a search space including four trivially 
different states. Then the next conjunctive goal, \texttt{(repeato x q)}, is 
applied to each of these states, producing four trivially different search 
spaces. In the examples above, the \conj{}s are allocating more computational 
resources to the search space of \texttt{a}. On the contrary, fair \conj{} 
would allocate resources evenly to each search space.

@@

A more interesting situation is when the first conjunct produces infinite many 
answers. Consider the following example, a naive specification of fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but no 
longer ones, which makes the strategies incomplete. A search strategy is 
\emph{complete} if and only if it can find out all the answers within finite 
time, otherwise, it is \emph{incomplete}.

@@

BFS puts states of the same cost in the same bag, and allocate resources 
carefully among search spaces related to different bags such that answers are 
produced in increasing order of cost. The \emph{cost} of a answer is its depth 
in the search tree (i.e. the number of calls to relational definitions required 
to find them) \citep{seres1999algebra}. In the following example, the cost of 
each answer is equal to the length of the inner lists plus the length of the 
outer list.

@@

\begin{defn}[fair \conj{}]
A \conj{} is fair if and only if it allocates computational resource evenly to 
search spaces produced from states in the same bag. A bag is a finite 
collection of state. And search strategies with fair \conj{} should represent 
search spaces with possibly infinite collection of state. 
\end{defn}

\begin{defn}[unfair \conj{}]
A \conj{} is unfair if and only if it is not fair.
\end{defn}

@@

Fig.~\ref{DFSi} shows part of the implementation of \DFSi{}. We follow a
convention to name variables bound to states with `s', to name variables bound 
to goals with `g', and to name variables bound to search spaces with a suffix 
`-inf'. The first function, \texttt{disj2}, implements binary disjunction. 
\texttt{append-inf} is its helper, which composes two disjunctive search 
spaces. 
The following function, \texttt{conj2}, implements binary conjunction. It 
applies 
the \emph{first} goal to the input state, then applies the second goal to 
states in the resulting search space. The latter process is done with a helper 
function. \texttt{append-map-inf} applies its input goal to states in its input 
search spaces and compose the resulting search spaces. It reuses 
\texttt{append-inf} for search space composition. The following three 
definitions introduce syntactic sugars that miniKanren users are more familiar 
with. The first two definitions say disjunction and disjunction are 
right-associative. The next and last definition says \conde{} relates its 
clauses disjunctively, and goals in the same clause conjunctively.

@@

\begin{figure}
	\lstinputlisting{Figures/balanced-disj.rkt}
	\caption{implementation of DFS$_\textrm{bi}$}
	\label{balanced-disj}
\end{figure}

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair 
\conj{}. The implementation of DFS$_\textrm{bi}$ differs from 
DFS$_\textrm{i}$'s 
in the \disj{} macro. We list the new \disj{} with its helper in 
Fig.~\ref{balanced-disj}. The helper \texttt{disj+} builds a balanced binary 
tree 
whose leaves are the goals and whose nodes are \texttt{disj2}s, hence the name 
of 
this search strategy. The first argument 
to \texttt{disj+} is a list of all goals. And the next two arguments accumulate 
goals in 
the left and right subtrees. The first clause says that when there is one goal, 
the tree is the goal itself. When there are more goals, the first argument is 
partitioned into two sub-lists. The partition is done by repetitively 
dispatching the first two goals (the last clause), until no goal remains (the 
second clause) or one goal remains (the third clause). In contrast, the \disj{} 
in DFS$_\textrm{i}$ constructs the binary tree with the same collection of leaf 
nodes but in a particularly unbalanced form.

@@

\begin{figure}
	\lstinputlisting{Figures/fDFS.rkt}
	\caption{implementation of DFS$_\textrm{f}$}
	\label{fDFS}
\end{figure}

@@

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our BFS implementation is not an 
exception: \texttt{append-map-inf/fair} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append-inf/fair} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus} respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s-inf} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some 
answers, \texttt{ks} is called with the first answer and the rest space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is like an eliminator of 
search space, hence the name. The remaining functions do the same thing as 
before.

@@

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than our BFS 
implementation. Details about 
difference in efficiency are in section 6.

@@

We devise two new search strategies (i.e. balanced interleaving DFS 
(DFS$_\textrm{bi}$) and fair DFS (DFS$_\textrm{f}$)) and devise a new 
implementation of BFS. These strategies have different features 
in fairness: $_\textrm{bi}$ has almost-fair \disj{} and unfair \conj{}. 
DFS$_\textrm{f}$ has fair \disj{} and unfair \conj{}. BFS has both fair \disj{} 
and fair \conj{}.
